---
title: "DAT 301 Lab 2B"
name: "Munachimso Aneke"
output:
  pdf_document: default
  html_document:
    df_print: paged
urlcolor: blue
---

The purpose of this lab is to explore data using the `dplyr` package from `tidyr`.  First, load the packages:

```{r}
library(dplyr)
library(tidyr)
```

The first portion of the lab will consider the British babynames in years 2017 and 1996.  The data below was obtained from the Office for National Statistics and compiled into a .CSV file in Excel.  Load the data using the below code and name the dataset `uknames_df`:

```{r}
uknames_df = read.csv("ukbabynames.csv", sep = ",", header = TRUE)
```


## Question 1

Create a dataframe that displays only the top 10 names in 2017 and 1996 for male and female babies born in the UK (suggestion: create four columns with each combination gender and year, example Boys2017, Girls2017,... ). Include commentary such as if you see any names appear on both lists or if the names are generally different, etc. 

```{r}
# Create slices of each group (Boys2017, Girls2017, Boys1996, Girls1996)
B2017 = uknames_df %>% filter(Gender == "M", Year == 2017) %>% 
  slice_max(Count, n = 10)

G2017 = uknames_df %>% filter(Gender == "F", Year == 2017) %>% 
  slice_max(Count, n = 10)

B1996 = uknames_df %>% filter(Gender == "M", Year == 1996) %>% 
  slice_max(Count, n = 10)

G1996 = uknames_df %>% filter(Gender == "F", Year == 1996) %>% 
  slice_max(Count, n = 10)

# extract the name colons for each group
Boys2017 = B2017[,"Name"]
Girls2017 = G2017[,"Name"]
Boys1996 = B1996[,"Name"]
Girls1996 = G1996[,"Name"]

Rank = c(1:10)

# create data.frame
TopTen = data.frame(Rank, Boys2017, Girls2017, Boys1996, Girls1996)
TopTen
```
Comparing the top ten names for girls in 1996 and 2017, the only name that repeats is "Emily". For boys names in 1996 and 2017, the only name that occurs in both years is "Jack". All other names in the top ten are distinct for both girls are boys.


## Question 2

### Part A

Create a dataframe with the counts of babies, separated by gender, and compare the results by using the top 10 names in the UK in the years 2017 and 1996. 

```{r}
Boys_2017 = sum(B2017[,"Count"])
Girls_2017 = sum(G2017[,"Count"])
Boys_1996 = sum(B1996[,"Count"])
Girls_1996 = sum(G1996[,"Count"])

TopTenCount = data.frame(Boys_2017, Girls_2017, Boys_1996, Girls_1996)
TopTenCount

```


### Part B

Create a dataframe that compares the proportion of babies, separated by gender, named top 10 names in the UK in the years 2017 and 1996

```{r}
boys2017 = Boys_2017/ sum(uknames_df %>% filter(Year == 2017, Gender == "M")%>%
                            select(Count))

girls2017 = Girls_2017/ sum(uknames_df%>%filter(Year == 2017, Gender == "F")%>%
                              select(Count))

boys1996 = Boys_1996/ sum(uknames_df%>%filter(Year == 1996, Gender == "M")%>%
                            select(Count))

girls1996 = Girls_1996/ sum(uknames_df%>%filter(Year == 1996, Gender == "F")%>%
                              select(Count))

TopTenProp = data.frame(boys2017, girls2017, boys1996, girls1996)
TopTenProp
```


### Part C

What trends do you see in parts A and B?

Part A shows the number of babies with top ten names in 1996 and 2017 by gender. The largest sum being the baby boys in 1996 and least sum being the baby girls in 2017. Part B shows the proportion of babies (by gender) in 1996 and 2017 that have a name from the top ten names. It appears that a larger proportion of boys in 1996 have a top ten name compared to boys in 2017. Similarly, a larger of girls in 1996  have a top ten name compared to girls in 2017.


## Question 3

Regardless of gender, what are the top 20 baby names in the UK in 2017 and what are their counts?

```{r}
UKTopTwenty2017 = uknames_df %>% filter(Year == 2017) %>% select(Name, Count)%>% 
  slice_max(Count, n=20)
UKTopTwenty2017
```



The next part of the analysis will cover US babynames.  The data obtained below is from the `babynames` package.  First load and save the dataset as `usnames_df`:

```{r}
library(babynames)
usnames_df = babynames
```



## Question 4

Create a column in the `usnames_df` dataset that displays the total number of babies named that specific baby name in the entire data set regardless of year.  Display the top 20 girl and top 20 boy names.

```{r}
# Top 20 Names for Girls 
TopGirls = usnames_df %>% filter(sex == "F") %>% group_by(name) %>% 
  summarise(count = sum(n))%>% slice_max(order_by = count, n = 20)
TopGirls

# Top 20 Names for Boys
TopBoys = usnames_df %>% filter(sex == "M") %>% group_by(name) %>% 
  summarise(count = sum(n))%>% slice_max(order_by = count, n = 20)
TopBoys

```



## Question 5

What percent of US girl and boy names were top 10 in 2017 vs. 1996?  Display answer in a dataframe.  Include commentary on anything you observe that you think is interesting.

```{r}
G1996_US = usnames_df %>% filter(sex == "F", year == 1996)%>% 
  slice_max(n, n = 10)
B1996_US = usnames_df %>% filter(sex == "M", year == 1996)%>% 
  slice_max(n, n = 10)
G2017_US = usnames_df %>% filter(sex == "F", year == 2017)%>% 
  slice_max(n, n = 10)
B2017_US = usnames_df %>% filter(sex == "M", year == 2017)%>% 
  slice_max(n, n = 10)


girls1996_US = sum(G1996_US$n)/ sum(usnames_df %>% filter(sex == "F", 
                                                     year == 1996)%>% select(n))
boys1996_US = sum(B1996_US$n)/ sum(usnames_df %>% filter(sex == "M", 
                                                     year == 1996)%>% select(n))
girls2017_US = sum(G2017_US$n)/ sum(usnames_df %>% filter(sex == "F", 
                                                     year == 2017)%>% select(n))
boys2017_US = sum(B2017_US$n)/ sum(usnames_df %>% filter(sex == "M", 
                                                     year == 2017)%>% select(n))

TopTenProp_US = data.frame(girls1996_US, boys1996_US, girls2017_US, boys2017_US)
TopTenProp_US

```
in 1996, a larger proportion of boys and girls were more likely to have a name in the top ten category compared to the boys and girls in 2017.


## Question 6

### Part A

Find the top 20 US names in 2017, regardless of gender.  Display with their counts.

```{r}
USTopTwenty2017 = usnames_df %>% filter(year == 2017) %>% select(name, n)%>% 
  slice_max(n, n=20)
USTopTwenty2017
```


### Part B

Compare the top 20 names in the UK and the US in 2017.  Which names were used in both the UK and the US?

```{r}
intersect(UKTopTwenty2017$Name, USTopTwenty2017$name)
```


## Question 7

Create a function that will look up a US babyname based on the name and gender and return the count of names of all babies in the dataset that have been named that name.

```{r}
# a_name - name to be searched
# a_sex - sex to be searched ("F" or "M")
countName = function(a_name, a_sex)
{
  df = usnames_df %>% group_by(name, sex) %>% summarise(count = sum(n))
  x = df$count[df$name == a_name & df$sex == a_sex]
  return(x)
}
```